---
output: html_document
editor_options: 
 chunk_output_type: console
---

# Compare PUF-Policy Engine and taxdata to IRS

-   pufpe -- PUF-Policy Engine
-   tctd -- Tax-Calculator / Taxdata
-   irs -- IRS aggregates

## Setup

```{r}
#| label: setup
#| output: false

source(here::here("r", "libraries.r"))
source(here::here("r", "constants.r"))
source(here::here("r", "functions.r"))

# pufpe <- open_dataset(dpqpufpe)

```


```{r}
#| label: get-tc-vars

tcvars <- readRDS(here::here("data", "tcvars.rds"))
# ht(tcvars)

utcvars <- tcvars |> 
  select(vname, vtype, desc) |> 
  distinct()

```


```{r}
#| label: get-targets

ptargets <- readRDS(here::here("data", "potential_targets.rds"))
# glimpse(ptargets)

# checkvals <- ptargets |> 
#   filter(year==2015) |> 
#   select(table, datatype, incsort, incrange, vname, irs=ptarget)

years <- c(2015, 2021)

irs_tab11 <- ptargets |> 
  filter(year %in% years, 
         table=="tab11", 
         datatype=="taxable",
         incsort > 1,
         vname %in% c("agi", "nret_all")) |> 
  select(year, incsort, incrange, vname, value=ptarget) |> 
  mutate(file="irs",
         vname=ifelse(vname=="nret_all", "nret", vname),
         value=ifelse(vname=="agi", value * 1000, value))

```



```{r}
#| label: get-tctd
#| output: false

# use this when we have year on the data files, but not until then
# pufpe <- open_dataset(
#   # sources = fs::path(dyfiles, "tax_microdata_2015.csv.gz"),
#   sources = dyfiles, 
#   col_types = schema(ISBN = string()),
#   format = "csv"
# )

tctd <- open_dataset(dpqtctd)

# tctd
# tctd |> glimpse()
# tctd |> ns()
# count(tctd, FLPDYR) |> collect()
# count(tctd, year) |> collect()
# 
# tctd |> 
#   summarise(n=n(), wt=sum(s006), wages=sum(e00200 * s006), .by=c(year)) |> 
#   arrange(year) |> 
#   collect()

```


```{r}
#| label: get-pufpe
#| output: false

# use this when we have year on the data files, but not until then
# pufpe <- open_dataset(
#   # sources = fs::path(dyfiles, "tax_microdata_2015.csv.gz"),
#   sources = dyfiles, 
#   col_types = schema(ISBN = string()),
#   format = "csv"
# )

pufpe <- open_dataset(dpqpufpe)

pufpe
pufpe |> glimpse()
pufpe |> ns()
count(pufpe, FLPDYR) |> collect()
count(pufpe, year) |> collect()

pufpe |> 
  summarise(n=n(), wt=sum(s006), wages=sum(e00200 * s006), .by=c(year, is_tax_filer)) |> 
  arrange(year) |> 
  collect()

```


## Prepare data for comparison to IRS spreadsheets for 2015 and 2021

Stack pufpe and tctd files for 2015 and 2021 and create a taxpayers file. The taxpayers file can be collapsed by AGI range and compared to different IRS spreadsheet totals.

```{r}
#| label: stack

years <- c(2015, 2021)

stack <- bind_rows(
   pufpe |>
     filter(year %in% years) |> 
     mutate(file="pufpe") |> 
     collect(),
   tctd |> 
     filter(year %in% years) |> 
     mutate(file="tctd") |> 
     collect()
  )

# ns(stack)

stack2 <- stack |> 
  mutate(irstottax = pmax(c09200 - refund, 0))

cat("Quantiles of calculated IRS total tax:\n")

stack2 |> 
  summarise(x=quantile(irstottax) |> list(), .by=file) |> 
  unnest_wider(col=x)

# new definition of taxpayers based on issue 37
# https://github.com/PSLmodels/tax-microdata-benchmarking/issues/37
taxpayers <- stack2 |> 
  filter(irstottax > 0) |> 
  select(file, year, s006, c00100, irstottax)

```


Define AGI cutpoints:

-   19 ranges for filers for all (?) tables, and for and taxpayers for Table 1.1
-   15 ranges for taxpayers for Table 1.2, Table 1.4, and  Table 2.1

```{r}
#| label: income-cuts

# what agi ranges are in the targets file?
# count(checkvals |> filter(table=="tab11"), vname)
# checkvals |> 
#   filter(table=="tab11", datatype=="filers", vname=="nret_all", incsort >1)


# define IRS income cutpoints for 19 income groups and

# use ycut19 for filers, all (?) IRS tables, and taxpayers tab11
ycut19 <- c(-Inf, 1, 
           seq(5e3, 30e3, 5e3),
           40e3, 50e3, 75e3, 100e3,
           200e3, 500e3,
           1e6, 1.5e6, 2e6, 5e6, 10e6, Inf)

# use this for taxpayers in tables other than tab11
ycut15 <- c(-Inf, 1, 
           seq(5e3, 30e3, 5e3),
           40e3, 50e3, 75e3, 100e3,
           200e3, 500e3,
           1e6, Inf)

```

Function to create comparison table

```{r}
#| label: tab-function

tab <- function(data, year, vname, title, scale=1, decimals=0){
  # create gt comparison table using a chosen variable
  
  tabdata <- data |> 
    filter(year==!!year, vname==!!vname) |> 
    select(incsort, incrange, file, value) |> 
    arrange(incsort) |> 
    pivot_wider(names_from = file) |> 
    select(incsort, incrange, irs, pufpe, tctd) |> 
    adorn_totals() |> 
    mutate(diff_pufpe=pufpe - irs,
           diff_tctd=tctd - irs,
           pdiff_pufpe=diff_pufpe / irs,
           pdiff_tctd=diff_tctd / irs)
  
  tab <- tabdata |>
    gt() |>
    tab_header(title = html(title),
               subtitle = html(paste0(vname, " -- ", year))) |>
  fmt_number(columns=-c(incsort, incrange, contains("pdiff")),
             scale=scale,
             decimals=decimals) |>
  fmt_percent(columns=contains("pdiff"),
             scale=1,
             decimals=1)
  tab
}

# tab(dflong, 2015, "nret")

```


## Comparison to IRS Table 1.1

Collapse data to appropriate income ranges.

```{r}
#| label: collapse-tab11
#| output: false

# tab11 uses ycut19 for filers and for taxpayers

stub <- irs_tab11 |> 
  select(year, vname, incsort, incrange) |> 
  distinct() |> 
  crossing(file=c("irs", "tctd", "pufpe"))
count(stub, year, file, incsort, incrange)

collapse <- taxpayers |>
  mutate(ycut=cut(c00100, ycut19, right=FALSE),
         incsort=as.integer(ycut) + 1L) |> # incsort will match with targets file
  summarise(nret = sum(s006), 
            agi = sum(s006*c00100),
            .by=c(file, year, incsort, ycut)) |>
  pivot_longer(cols=c(nret, agi), names_to = "vname")

dflong <- stub |> 
  filter(file!="irs") |> 
  left_join(collapse |> select(-ycut), join_by(year, vname, incsort, file)) |> 
  bind_rows(irs_tab11)

glimpse(dflong)
count(dflong, incsort, incrange)
count(dflong, vname, year, file)


```


## Number of returns 2015 and 2021

```{r}

tab(dflong, 2015, "nret", title="Number of taxable returns (thousands)", scale=1e-3, decimals=1)

```


```{r}

tab(dflong, 2021, "nret", title="Number of taxable returns (thousands)", scale=1e-3, decimals=1)

```


## AGI 2015 and 2021


```{r}

tab(dflong, 2015, "agi", title="Adjusted gross income ($ billions), taxable returns", scale=1e-9, decimals=1)

```


```{r}

tab(dflong, 2021, "agi", title="Adjusted gross income ($ billions), taxable returns", scale=1e-9, decimals=1)

```



