---
output: html_document
editor_options: 
 chunk_output_type: console
---

# Find Variables in PUF-PolicyEngine Data That Have Weighted Sums of Zero

I examine 2015 because it allows me to compare to the raw PUF, from which the PUF-PolicyEngine draws its tax filers. If variables are zero in PUF-PolicyEngine but nonzero in the PUF, However, inspection shows that zero-valued variables in 2015 are also zero-valued in other years.

Compare to other files:

-   pufpe -- PUF-Policy Engine vintage 4/10/2024, which is a 2023 hybrid file based on CPS and puf.csv, backcasted to 2015
-   tctd -- Tax-Calculator / taxdata vintage 4/5/2024 -- based on puf.csv from taxdata, grown to 2015 with default Tax-Calculator assumptions and methods
-   puf2015 -- Raw puf for 2015 (as provided by IRS SOI)

## Setup

```{r}
#| label: setup
#| output: false

source(here::here("r", "libraries.r"))
source(here::here("r", "constants.r"))
source(here::here("r", "functions.r"))

```

## Get various data files

```{r}
#| label: get-variable-labels
#| output: false

tcvars <- readRDS(here::here("data", "tcvars.rds"))
# ht(tcvars)

utcvars <- tcvars |> 
  select(vname, vtype, desc) |> 
  distinct()

count(utcvars, vtype)

invars <- utcvars |> 
  filter(vtype=="read") |> 
  pull(vname)

```


```{r}
#| label: get-puf2015
#| output: false

pufdir <- r"(E:\data\puf_files\puf2015)"
ppath <- fs::path(pufdir, "puf_2015.csv")

puf <- vroom(ppath)
puf2 <- puf |> 
  lcnames() |> 
  mutate(year=2015, s006=s006 / 100)

glimpse(puf2)
# summary(puf2)
# skim(puf2)

```


```{r}
#| label: get-stack-data
#| output: false

tctd <- open_dataset(dpqtctd)
pufpe <- open_dataset(dpqpufpe)

# quick check on e26270
# tctd |> 
#   select(year, s006, e26270) |> 
#   summarise(n=n(), wtdn=sum(s006), e26270=sum(e26270 * s006), .by=year) |> 
#   collect()
# 
# pufpe |> 
#   select(year, s006, e26270) |> 
#   summarise(n=n(), wtdn=sum(s006), e26270=sum(e26270 * s006), .by=year) |> 
#   collect()

ykeep <- 2015

stack <- bind_rows(
  pufpe |> 
    filter(year==ykeep) |> 
    select(year, any_of(invars)) |> 
    mutate(file="pufpe") |> 
    collect(),
  tctd |> 
    filter(year==ykeep) |> 
    select(year, any_of(invars)) |> 
    mutate(file="tctd") |> 
    collect(),
  puf2 |> 
    select(year, any_of(invars)) |> 
    mutate(file="puf2015"))

glimpse(stack)

```


```{r}
#| label: get-weighted-sums
#| output: false

wsums <- stack |> 
  select(file, year, s006, starts_with("e", ignore.case=FALSE), starts_with("p", ignore.case=FALSE)) |> 
  pivot_longer(-c(file, year, s006)) |> 
  filter(!str_detect(name, "_")) |> 
  summarise(wtdsum=sum(value * s006), .by=c(file, year, name)) |> 
  pivot_wider(names_from = file, values_from = wtdsum)
wsums

```

## Show variables where PUF-PolicyEngine weighted sums are zero

```{r}
#| label: show-zero-values

wsums2 <- wsums |> 
  left_join(utcvars |> 
              select(name=vname, desc),
            by = join_by(name)) |> 
  mutate(adiff=abs(pufpe - puf2015)) |> 
  arrange(desc(adiff))

wsums3 <- wsums2 |> 
  filter(pufpe==0) |>
  mutate(rank=row_number()) |> 
  select(rank, name, tctd, puf2015, pufpe, adiff, desc)

wsums3 |> 
  gt() |> 
  tab_header(title = html("'e' and 'p' variables that have zero values in the 2015 PUF-PolicyEngine file"),
             subtitle="Compared to raw 2015 PUF and 2015 taxdata file, $ billions, sorted by descending absolute (pufpe - puf2015)") |> 
  fmt_number(columns = -c(rank, name, desc),
             scale=1e-9,
             decimals=1) 

```

## Check zero-valued PUF-PolicyEngine variables in all years

Do the PUF-PolicyEngine variables that have zero weighted-sums in 2015 have zero weighted-sums in all years?

```{r}
#| label: all-years

zvars <- wsums3$name

pufpe |>
  select(year, s006, all_of(zvars)) |>
  group_by(year) |> 
  summarise(across(-s006, ~sum(.x * s006))) |>
  ungroup() |> 
  collect() |> 
  pivot_longer(-year) |> 
  pivot_wider(names_from = year) |> 
  gt() |> 
  tab_header(title = html("Weighted sums in the PUF-PolicyEngine file each year"),
             subtitle=html("'e' and 'p' variables where weighted sums are zero in the 2015 PUF-PolicyEngine file"))


```


